# Data Structures

## Objectives

- Learn how to construct vectors, data frames, matrices, arrays, and lists
- Learn how each data structure relates to the others
- Subset and modify objects  
- Know when to use each structure 

## Vectors

Vectors are the basic building block of R.  
They contain elements of the same type.

```{r}
# Character
x <- c("a", "b", "c")

# Numeric (default number type)
y <- c(1, 2, 3)

# Integer
z <- c(1L, 2L, 3L)

# Logical
w <- c(TRUE, FALSE, TRUE)

class(y)
```

### Subsetting

Use brackets `[]`.

```{r}
x <- c("a", "b", "c", "d")

x[1]        # first element
x[2:4]      # range
x[c(1,4)]   # multiple positions
```

Logical subsetting:

```{r}
y <- 1:5
y[y > 3]
```

Replace values:

```{r}
y[y > 3] <- 0
y
```

### Factors

Factors represent categorical data.

```{r}
names <- c("John", "Mary", "John", "Mary")
f <- factor(names)

f
levels(f)
table(f)
```

---

## Data Frames

A data frame is a 2D structure (rows × columns).  
Each column is a vector.  
Columns may have different types.

```{r}
df <- data.frame(
  name = c("Ana", "Ben", "Chris"),
  age  = c(20, 21, 19),
  passed = c(TRUE, TRUE, FALSE)
)

str(df)
```

### Loading and Saving

Read CSV:

```r
df <- read.csv("data.csv")
```

Write CSV:

```r
write.csv(df, "output.csv", row.names = FALSE)
```

### Subsetting and Combining

Extract column:

```{r}
df$age
df[, "age"]
```

Extract rows:

```{r}
df[df$age > 19, ]
```

Add column:

```{r}
df$status <- df$age > 20
```

Combine:

```{r}
rbind(df, df)
cbind(df, new_col = 1:3)
```

---

## Matrices and Arrays

A matrix is a vector with dimensions.  
All elements must be the same type.

```{r}
m <- matrix(1:12, nrow = 3, ncol = 4)

m[2,3]   # row 2, column 3
m[2, ]   # row 2
m[, 4]   # column 4
```

Matrix operations:

```{r}
t(m)          # transpose
m * m         # element-wise
m %*% t(m)    # matrix multiplication
```

An array is a multi-dimensional matrix:

```{r}
a <- array(1:8, dim = c(2,2,2))
```

---

## Lists

A list can store different types and sizes of objects.

```{r}
my_list <- list(
  numbers = 1:3,
  name = "Ana",
  matrix = matrix(1:4, 2),
  df = df
)

my_list
```

Subsetting lists:

```{r}
my_list[1]      # returns sub-list
my_list[[1]]    # returns element
my_list$name
```

## Summary: When to Use What

| Structure | Same Type? | Dimensions | Use Case |
|------------|------------|------------|-----------|
| Vector | Yes | 1D | Single variable |
| Factor | Yes | 1D | Categories |
| Data Frame | No (by column) | 2D | Tabular data |
| Matrix | Yes | 2D | Math operations |
| Array | Yes | 3D+ | Multidimensional data |
| List | No | Any | Complex objects |

Core idea: Everything in R builds from vectors.