# External Packages

## Installing Packages

One of the biggest strengths of R is its large ecosystem of **packages** (also called libraries). Packages extend R by adding new functions for data analysis, statistics, visualization, and machine learning. Thousands of packages are available through the **Comprehensive R Archive Network (CRAN)**.

To use a package, you first install it (one time) and then load it into your session.

```r
install.packages("ggplot2")
library(ggplot2)
```

Once loaded, the functions from the package become available for use.

Packages are commonly used to:

- Perform statistical modeling
- Manipulate and analyze datasets
- Create high-quality visualizations
- Extend R with specialized tools

## Built-in Statistical Functions

R already includes many powerful statistical tools in its **base and stats packages**, which are loaded automatically when R starts.

For example, you can run statistical tests or fit models directly.

### Linear Regression

The `lm()` function fits a linear model.

```{r}
model <- lm(mpg ~ wt, data = mtcars)
summary(model)
```

This model estimates how **vehicle weight (`wt`) affects fuel efficiency (`mpg`)**.

### t-tests

You can also perform hypothesis testing using functions such as `t.test()`.

Example: Compare the mean miles-per-gallon for cars with different transmission types.

```{r}
t.test(mpg ~ am, data = mtcars)
```

These built-in statistical tools make R a powerful environment for quantitative analysis.

## Example: Creating Visualizations with ggplot2

The **ggplot2** package is one of the most widely used tools for creating data visualizations in R. It implements the *Grammar of Graphics*, which builds plots layer by layer.

For example, we can visualize the relationship between engine weight and fuel efficiency using the built-in `mtcars` dataset.

```{r}
library(ggplot2)

ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point()
```

This creates a scatter plot where:

- `wt` (vehicle weight) is on the x-axis
- `mpg` (miles per gallon) is on the y-axis

Additional layers can be added to customize the plot.

```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(
    title = "Fuel Efficiency vs Vehicle Weight",
    x = "Weight",
    y = "Miles per Gallon"
  )
```

## Looking Ahead

While base R already includes powerful statistical tools, many analysts extend their workflows with additional packages. Some commonly used packages include:

- `ggplot2` – data visualization  
- `dplyr` – data manipulation  
- `tidyr` – reshaping data  

Together these packages form part of the **tidyverse**, a widely used ecosystem for data science in R.

In more advanced courses and projects, you will learn how to combine statistical modeling, visualization, and data manipulation to build complete data analysis workflows.
