# Functions

Functions are reusable blocks of code designed to perform a specific task.  
Writing functions improves clarity, efficiency, and reproducibility.

## Objectives

- Define and use built-in functions  
- Write custom functions  
- Return single and multiple values  
- Use default arguments  
- Create nested functions  
- Validate arguments using `stopifnot()`  
- Understand basic error handling  

## Built-In Functions

R provides many built-in functions for analysis, including:

- `mean()`
- `min()`
- `max()`
- `summary()`
- `quantile()`

### Example: mean()

```{r}
mean(1:5)
```

Handling missing values:

```{r}
v <- c(2, NA, 4, NaN, 6)

mean(v)                # returns NA
mean(v, na.rm = TRUE)  # removes missing values
```

To use functions from packages:

```r
library(dplyr)

# Example of package function
dplyr::select
```

## Custom Functions

You create a function using `function()`.

### Basic Syntax

```r
f <- function(arg1, arg2, arg3, ...){
  # code
}
```

### Example: One Argument

```{r}
squareroot <- function(a){
  a^0.5
}

squareroot(49)
```

### Example: Two Arguments

```{r}
Addtwo <- function(a, b){
  a + b
}

Addtwo(1, 2)
```

### Returning Values

R automatically returns the last evaluated expression, but `return()` makes it explicit.

```{r}
F2C <- function(temp) {
  c <- (temp - 32) * (5 / 9)
  return(c)
}

F2C(100)
```

### Returning Multiple Values

Use a list to return multiple outputs.

```r
sqsum <- function(a){
  sq <- a^0.5
  sumsq <- a + sq
  output <- list(sq = sq, sumsq = sumsq)
  return(output)
}

sqsum(49)
```



## Default Arguments

You can assign default values to arguments.

```r
power <- function(x, exponent = 2){
  x^exponent
}

power(3)       # uses default exponent = 2
power(3, 3)    # overrides default
```

Default arguments make functions more flexible.



## Nested Functions

Nested functions combine multiple operations inside one expression.

Example using `mtcars`:

Goal: Compute the mean `mpg` for cars with 4 cylinders.

Step-by-step approach:

```r
data(mtcars)

ind1 <- mtcars$cyl == 4
ind2 <- mtcars$mpg[ind1]
mean(ind2)
```

Nested version:

```r
mean(mtcars$mpg[mtcars$cyl == 4])
```

You can also define functions inside functions:

```r
outer_function <- function(x){
  
  inner_function <- function(y){
    y^2
  }
  
  inner_function(x) + 1
}

outer_function(3)
```



## Error Handling

Functions should check inputs to avoid incorrect results.

### Using `stopifnot()`

```r
safe_sqrt <- function(x){
  stopifnot(x >= 0)
  sqrt(x)
}

safe_sqrt(9)
```

If the condition fails, R stops execution.



### Using `stop()`

```r
safe_divide <- function(a, b){
  if(b == 0){
    stop("Division by zero is not allowed.")
  }
  a / b
}
```

### Using `warning()`

```r
check_positive <- function(x){
  if(x < 0){
    warning("Negative value detected.")
  }
  x
}
```

## Key Points

- Functions improve efficiency and readability  
- Built-in functions cover many common tasks  
- Custom functions use `function()`  
- Use default arguments for flexibility  
- Return multiple values using lists  
- Nested functions simplify complex operations  
- Validate inputs using `stopifnot()` or `stop()`  
