# If-Else Statements

## Objectives

- Write conditional statements using `if`
- Use `if...else` and `else if`
- Apply vectorized logic using `ifelse()`
- Understand when to use each structure

## If Statements

Basic syntax:

```r
if (condition){
  # do task
}
```

Example:

```{r}
a <- 5

if (a > 3){
  print("a is bigger than 3")
}
```

The code inside the braces runs only if the condition is TRUE.

## Else Statements

Use `else` to define what happens when the condition is FALSE.

```{r}
a <- 5

if (a > 3){
  print("a is bigger than 3")
} else {
  print("a is NOT bigger than 3")
}
```

Only one block will execute.

## Else-If Statements

Use `else if` to test multiple conditions.

```{r}
a <- 5

if (a > 3){
  print("a is bigger than 3")
} else if (a == 3) {
  print("a equals 3")
} else {
  print("a is less than 3")
}
```

R checks conditions from top to bottom and stops when one is TRUE.

## The `ifelse()` Function

`ifelse()` is a vectorized version of `if`.

Syntax:

```r
ifelse(condition, value_if_true, value_if_false)
```

Example:

```{r}
a <- 5

ifelse(a > 3,
       "a is bigger than 3",
       "a is not bigger than 3")
```

Unlike `if`, `ifelse()` works well on vectors:

```{r}
x <- 1:5
ifelse(x > 3, "big", "small")
```

Use `ifelse()` when working with vectors or data frame columns.
