# Loops

## Objectives

- Write and understand `for` loops
- Write and understand `while` loops
- Use `apply` family functions as loop alternatives
- Apply functions over lists and matrices
- Understand `apply`, `lapply`, `sapply`, `mapply`, and `tapply`

## For Loops

Full syntax:

```r
for (iterator in sequence){
  # do task
}
```

Example:

```{r}
for (i in 1:5){
  print(i)
}
```

Using `seq()`:

```{r}
for (i in seq(1, 5)){
  print(letters[i])
}
```

Short form:

```{r}
for (i in 1:5) print(i)
```

Use `for` loops when you know how many iterations you need.

## While Loops

Syntax:

```r
while (condition){
  # do task
}
```

Example:

```{r}
a <- 1

while (a < 5){
  print(a)
  a <- a + 1
}
```

The loop continues as long as the condition is TRUE.

Be careful: if the condition never becomes FALSE, the loop runs forever.

## Apply Functions

The `apply` family provides cleaner alternatives to explicit loops.

There are functions in R that make looping easier:

- `apply()` – apply a function over margins of an array  
- `lapply()` – loop over a list and return a list  
- `sapply()` – simplified version of `lapply()`  
- `mapply()` – multivariate version of `lapply()`  
- `tapply()` – apply a function over subsets of a vector  

### apply()

Most often used to apply a function to rows or columns of a matrix.

```{r}
m <- matrix(1:12, 3, 4)
print(m)

apply(m, 1, sum)  # row sums
apply(m, 2, sum)  # column sums
```

Syntax:

```r
apply(X, MARGIN, FUN)
```

- `X` = matrix or array  
- `MARGIN = 1` (rows) or `2` (columns)  
- `FUN` = function to apply  

Similar built-in matrix functions:

```{r}
rowSums(m)
colSums(m)
rowMeans(m)
colMeans(m)
```

Note: `apply()` is not necessarily faster than loops, but it simplifies code.

### lapply() and sapply()

l = long (returns list)  
s = short (simplifies output)

```{r}
list1 <- list(
  l1 = seq(1, 10),
  l2 = 20:29,
  l3 = rnorm(4)
)

lapply(list1, mean)
sapply(list1, mean)
```

- `lapply()` always returns a list  
- `sapply()` tries to simplify the result (vector or matrix)

### mapply()

Multivariate version of `lapply()`.

```{r}
mapply(sum, 1:3, 4:6)
```

Applies a function to multiple arguments element-wise.

### tapply()

Applies a function over subsets of a vector.

```{r}
x <- 1:10
group <- rep(c("A", "B"), each = 5)

tapply(x, group, mean)
```

Useful for grouped summaries.

## Key Points

- Use `for` and `while` for explicit iteration  
- Use `apply` family functions for cleaner, more readable code  
- `lapply()` returns lists  
- `sapply()` simplifies output  
- `apply()` works well on matrices  

